---
title: "infectionAgeHIV Workflow"
author: "Erik Lundgren"
date: "July 2023"
output:
  html_document: default
  pdf_document: default
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

**infectionAgeHIV** is a small package that can produce probability distributions for infection times based on HIV biomarkers

## Workflow

- format patient data into two data frames: one for data where there is one value per patient and one where there may be multiple
- use `find.infection.ages` to determine which biomarker values are usable in the multiple biomarker model and find infection age distributions

## Load package and example data and set pRNG seed

```{r load}
library(infectionAgeHIV)
df.single <- df.single.example
df.multiple <- df.multiple.example
set.seed(12345)
```

## Show format of example data
```{r }
#truncate sequences to be able to be displayed on page
df.multiple.substr <- df.multiple.example
df.multiple.substr$value[c(3,7,9,11)] <- paste0(substr(df.multiple.substr$value[c(3,7,9,11)], 1, 30), "...")
knitr::kable(df.single.example)
knitr::kable(df.multiple.substr)
```

## Using biomarker data to predict infection times

```{r run_mbm}
df <- find.infection.ages(df.single = df.single,
                          df.multiple = df.multiple, 
                          n.adapt = 1e4, n.burn = 1e5, n.iter = 1e6)
```

We can plot the inferred distributions for the amount of time between infection and diagnosis.

```{r plot_diag}
plot(df$infection_age_dists_diag[[1]], 
     xlab = "Time between infection and diagnosis (years)",
     main = paste0("Patient ", df.single$patient_id[[1]]))
plot(df$infection_age_dists_diag[[2]], 
     xlab = "Time between infection and diagnosis (years)",
     main = paste0("Patient ", df.single$patient_id[[2]]))
```

The distributions for the amount of time between infection and taking sequence samples 
and the infection time density in terms of real time are also returned.

```{r plot_seq}
plot(df$infection_age_dists_seq[[2]], 
     xlab = "Time between infection and first sequence sampling (years)",
     main = paste0("Patient ", df.single$patient_id[[2]]))
plot(df$pdf[[2]], type = "l",
     xlab = "Infection Time",
     ylab = "Density",
     main = paste0("Patient ", df.single$patient_id[[2]]))
```

## Drawing random infection ages

We can also draw from the inferred distributions by supplying a uniform [0,1] random variable
to the inverse CDF transform.

```{r draw_times}
drawn.infection.ages <- df$icdf[[2]](runif(10))
print(drawn.infection.ages)
```

## Using `prepare.HIV.data`

If more control is needed, `prepare.HIV.data` can be called without the wrapper `find.infection.ages`